Oriented Gradient Histogram applied to P300 Detection

### 2017-12-07 article.tex
cambié la figura del keypoint
últimas modificaciones

\begin{figure}[H]
\centering
\subfigure[Plot of the signal, a keypoint and the corresponding patch.]{\includegraphics[width=.49\linewidth]{DiagramaConkeypoint.png}\label{patch}}
\subfigure[Orientations on two blocks of the patch.]{\includegraphics[width=.49\linewidth]{histogramsbines.png}\label{histogrambines}}
%\includegraphics[width=9cm]{gradientszoom.png}
\caption{ Example of a patch and a scheme of the orientation's histogram computation.}
\label{fig:sampledescriptor}
\end{figure}

\subsubsection{Classification}
\label{Classification}

The following step is classify the descriptors $\{d_{kp}, kp =1,\dots, 12\}$ in order to determine the correct row and column of the chosen letter on the Speller Matrix.   The classification is carried out by using a discriminative semi-supervised classification method, based on the Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008}.

@@

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item Obtain the descriptor from the image of the signal plot according to the method described in Section~\ref{SIFT}, $ d^r_1, \dots,  d^r_6 $ for rows and  $ d^c_7, \dots,  d^c_{12} $ for columns. 
\item Compute

@@

where $NN_T(d^l_u), \;l=r,c$ is the set of the $k$ nearest neighbors to $d^l_u, \;l=r,c$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l=r,c$.  By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{r} $ and column $ \hat{c} $. The parameter $k$ is an input parameter of the algorithm.

%The first step consists in extracting labeled \textit{hit} descriptors from the training set which are thus grouped together in a KD-tree \citep{Vedaldi2010} template set $ T $.  On decoding stage, 12 new images from the signal segments are generated, and their descriptors extracted.  The 12 descriptors are divided in rows and columns.  For each one of the 6 query descriptors $ q $ for row/column, its $ k $ nearest neighbours in the template set $T$, $ NN_T^k(q) $ (acquired during the training phase)  are obtained and the distance between each one of them, $ x_i $, and the query descriptor $ q $ is summarized.  The query descriptor $ q $ that minimizes this summation is the one that is chosen, first among the 6 rows, and then among the remaining 6 columns.  By performing this procedure, the correct letter can be identified by matching the row $ r $ and column $ c $ from the P300 speller matrix.

@@

%Este párrafo no se entiende nada
%------------------
Especially in the case of the P300 response, the oddball paradigm requires that one of the stimuli needs to be infrequent. Hence this will unavoidably force the data to be unbalanced~\citep{Tibon2015}.  At the same time, the NBNN method suffers from biased classification on unbalanced classes~\citep{Fornoni2014}.  By reversing the roles of the query and the class in equations \ref{eq:multiclassificationrow} and \ref{eq:multiclassificationcol}, it is only necessary to obtain the template set $ T $ with the learned descriptors representative of the P300 ERP, hence avoiding the problem of unbalanced classes. Figure~\ref{fig:classification} shows a schema of this idea,  where the classification is only performed against the template set T, for each one of the twelve classes, thus avoiding the unbalanced situation (2 hit vs. 10 no-hit).
\begin{figure}[H]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{For each letter, single trial segments are averaged for the 6 rows and 6 columns.}
\label{fig:classification}
\end{figure}
%------------------
\subsection{Experimental Protocol} \label{Protocol}

To verify the validity of the proposed framework and method, the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website~\citep{Brunner2014} by  IRCCS Fondazione Santa Lucia, was used to perform an offline BCI Simulation to decode the spelled words from the provided signals.  The algorithm was implemented using  VLFeat~\citep{Vedaldi2010} Computer Vision libraries on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).

@@

The experimental protocol used to generate this dataset is explained in~\citep{Riccio2013} but can be summarized as follows:  8 subjects with confirmed diagnoses but on different stages of ALS disease, were recruited and accepted to perform the experiments. The P300 detection task designed for this experiment consisted of spelling 7 words of 5 letters each, using the traditional P300 Speller Matrix~\citep{Farwell1988}. The flashing of rows and columns provide the deviant stimulus required to elicit this physiological response.  The first 3 words are used for training and the remaining 4 words, for testing with visual feedback.  A trial, as defined by the BCI2000 platform~\citep{Schalk2004}, is every attempt to select a letter from the speller. It is composed of signal segments corresponding to 10 repetitions of flashes of 6 rows and 10 repetitions of flashes of 6 columns of the matrix, yielding 120 repetitions.  Flashing of a row or a column is performed for 0.125 s, following by a resting period (i.e. Inter-stimulus interval) of the same length.  After 120 repetitions an inter-trial pause is included before resuming with the following letter.

The recorded dataset was sampled at 256 Hz and consisted of an EEG matrix for electrode channels Fz,Cz,Pz,Oz,P3,P4,PO7 and PO8, identified according to the 10-20 International System,  for each one of the 8 subjects.

%Este párrafo no se entiende nada y repite in order
In order to asses and verify the identification of the P300 response, subjects are instructed to perform a copy-spelling task. They have to fix their attention to successive letters in order to copy a previously determined set of words (in contrast to an online free-running operation of the speller where each users decides on its own what letter to choose).

@@

We replicated the same experiment on healthy subjects using g.Tec g.Nautilus 8 channels.  The experiment protocol is exactly the same and we 20 subjects, age $ 30$, 5 females and 15 males, 6 right handed and the remaining left handed.

Participants were recruited voluntarily and the experiment were conducted anonymyzing information in accordance with the declaration of Helsinki.  No monetary compensation was delivered and all participants agreed and signed a written informed consent.  All healthy participants (n=13, 7 males, 6 females, average age 22.5 years (SD = 1,8 years), range 21–26 years) had normal or corrected-to normal vision and no history of neuropsychiatric dis- orders.

EEG data were recorded from the same 8 electrodes positions (C3, Cz, C4, CPz, P3, Pz, P4, POz) with all the systems. The reference positions were placed at the right ear. The ground positions was placed at the left ear, except with the g.Nautilus system (placed at AFz, because this could not be changed). Sampling rate was set at 250 Hz, the closest possible to predefined g.Mobilab sampling rate (256 Hz). Data were acquired notch filtered at 50 Hz and passband filtered between 2 Hz and 30 Hz. The filtering was performed via the proprietary Simulink HighSpeed Online Processing block modules for g.MOBIlab and g.Nautilus systems, and via MATLAB code for the Xpress system. Table 1 summarizes the acquisition settings for each system.

8 gel-based active electrodes (g.LADYbird) + g.LADYbird (GND) + g.GAMMAearclip (REF) C3, Cz, C4, CPz, P3, Pz, P4, POz, GND: AFz, REF: right ear

>>>>

\begin{figure}[H]
\centering
\subfigure[Plot of the signal, a keypoint and the corresponding patch.]{\includegraphics[width=.49\linewidth]{signalpatchkeypoint.png}\label{patch}}
\subfigure[Orientations on two blocks of the patch.]{\includegraphics[width=.49\linewidth]{histogramsbines.png}\label{histogrambines}}
%\includegraphics[width=9cm]{gradientszoom.png}
\caption{ Example of a patch and a scheme of the orientation's histogram computation.}
\label{fig:sampledescriptor}
\end{figure}

\subsubsection{Speller Matrix letter Identification}
\label{Classification}

%The following step is classify the descriptors $\{d_{kp}, kp =1,\dots, 12\}$ in order to determine the correct row and column of the chosen letter on the Speller Matrix.   The classification is carried out by using a discriminative semi-supervised classification method, based on the Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008}.

@@

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item Repeat item 1 $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t$ and $c$ correspond to time and chanel, respectively. The parameter $k_a$ is an input of the algorithm.
\item Compute  
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c)
\label{averaging}
\end{equation}

\item Plot the signal $x(t,c)$, according Section~\ref{Plot}.

\item Obtain the descriptor from the image of the signal plot of $x(t,c)$ according to the method described in Section~\ref{SIFT}, $ d^r_1, \dots,  d^r_6 $ for rows and  $ d^c_7, \dots,  d^c_{12} $ for columns.

\item Match with the Template $T$ by computing

@@

where $NN_T(d^l_u), \;l=r,c$ is the set of the $k$ nearest neighbors to $d^l_u, \;l=r,c$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l=r,c$.  By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{r} $ and column $ \hat{c} $. The  $k$ value is an input parameter of the algorithm.

% By reversing the roles of the query and the class in equations \ref{eq:multiclassificationrow} and \ref{eq:multiclassificationcol}, it is only necessary to obtain the template set $ T $ with the learned descriptors representative of the P300 ERP, hence avoiding the problem of unbalanced classes. 
%The first step consists in extracting labeled \textit{hit} descriptors from the training set which are thus grouped together in a KD-tree \citep{Vedaldi2010} template set $ T $.  On decoding stage, 12 new images from the signal segments are generated, and their descriptors extracted.  The 12 descriptors are divided in rows and columns.  For each one of the 6 query descriptors $ q $ for row/column, its $ k $ nearest neighbours in the template set $T$, $ NN_T^k(q) $ (acquired during the training phase)  are obtained and the distance between each one of them, $ x_i $, and the query descriptor $ q $ is summarized.  The query descriptor $ q $ that minimizes this summation is the one that is chosen, first among the 6 rows, and then among the remaining 6 columns.  By performing this procedure, the correct letter can be identified by matching the row $ r $ and column $ c $ from the P300 speller matrix.

@@

Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments are averaged for the 6 rows and 6 columns.}
\label{fig:classification}
\end{figure}

%In the case of the P300 response, the oddball paradigm requires that one of the stimuli be infrequent. Hence this forces the data to be unbalanced~\citep{Tibon2015}.  At the same time, the NBNN method suffers from biased classification on unbalanced classes~\citep{Fornoni2014}. %Para solucionar este problema,

%Este párrafo no se entiende nada
%------------------

%------------------
\subsection{Experimental Protocol} \label{Protocol}

To verify the validity of the proposed framework and method, the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website~\citep{Brunner2014} by  IRCCS Fondazione Santa Lucia, is used. An own dataset with  the same experimental conditions is generated. Both of them are utilized to perform an offline BCI Simulation to decode the spelled words from the provided signals.

The algorithm was implemented using  VLFeat~\citep{Vedaldi2010} Computer Vision libraries on MATLAB V2014a (Mathworks Inc., Natick, MA, USA).

In the following sections the characteristics of the datasets and parameters of the identification algorithm are described.

@@

The experimental protocol used to generate this dataset is explained in~\citep{Riccio2013} but can be summarized as follows:  8 subjects with confirmed diagnoses but on different stages of ALS disease, were recruited and accepted to perform the experiments. The P300 detection task designed for this experiment consisted of spelling 7 words of 5 letters each, using the traditional P300 Speller Matrix~\citep{Farwell1988}. The flashing of rows and columns provide the deviant stimulus required to elicit this physiological response.  The first 3 words are used for training and the remaining 4 words, for testing with visual feedback.  A trial, as defined by the BCI2000 platform~\citep{Schalk2004}, is every attempt to select a letter from the speller. It is composed of signal segments corresponding to $k_a =10$ repetitions of flashes of 6 rows and $k_a =10$ repetitions of flashes of 6 columns of the matrix, yielding 120 repetitions.  Flashing of a row or a column is performed for 0.125 s, following by a resting period (i.e. Inter-stimulus interval) of the same length.  After 120 repetitions an inter-trial pause is included before resuming with the following letter.

The recorded dataset was sampled at 256 Hz and it consisted of an EEG matrix for electrode channels Fz,Cz,Pz,Oz,P3,P4,PO7 and PO8, identified according to the 10-20 International System,  for each one of the 8 subjects.

%Este párrafo no se entiende nada y repite in order
In order to asses and verify the identification of the P300 response, subjects are instructed to perform a copy-spelling task. They have to fix their attention to successive letters for copying a previously determined set of words, in contrast to a free-running operation of the speller where each user decides on its own what letter to choose.

@@

We replicat the experiment on healthy subjects using g.Tec g.Nautilus 8 channels.  The experiment protocol is  the same and we 20 subjects, age $ 30$, 5 females and 15 males, 6 right handed and the remaining left handed.

Participants were recruited voluntarily and the experiment were conducted under anonymous information in accordance with the declaration of Helsinki.  No monetary compensation was delivered and all participants agreed and signed a written informed consent.  All healthy participants had normal or corrected-to normal vision and no history of neuropsychiatric disorders. The experiment was performed with 13 subjects, 7 males, 6 females, average age 22.5 years, standard deviation  1,8 years, range 21–26 years.

EEG data were recorded from the same 8 electrodes positions (C3, Cz, C4, CPz, P3, Pz, P4, POz) with all the systems. The reference positions were placed at the right ear. The ground positions was placed at the left ear, except with the g.Nautilus system (placed at AFz, because this could not be changed). Sampling rate was set at 250 Hz, the closest possible to predefined g.Mobilab sampling rate (256 Hz). Data were acquired notch filtered at 50 Hz and passband filtered between 2 Hz and 30 Hz. The filtering was performed via the proprietary Simulink HighSpeed Online Processing block modules for g.MOBIlab and g.Nautilus systems, and via MATLAB code for the Xpress system. Table~\ref{tab:results} summarizes the acquisition settings for each system.

%8 gel-based active electrodes (g.LADYbird) + g.LADYbird (GND) + g.GAMMAearclip (REF) C3, Cz, C4, CPz, P3, Pz, P4, POz, GND: AFz, REF: right ear
